---
title: R Review & Assessment
output: learnr::tutorial
runtime: shiny_prerendered
time_slot: 60 minutes"
---


```{r, include = FALSE}
library(learnr)
```

## Introduction

Let's take about an hour to take a quick assessment, touch up on basic R programming skills in a survey style, and then we'll move onto some challenges which should test your understanding if we have time.

## Learning Outcomes

- Assess where everyone is with R so we can shape the curricula
- Refresh general R programming skills
- Test R skills against some example problems

## Assessment

Everyone will arrive to this lesson with different experiences with R.
Skill with R doesn't necessarily exist a continuum and can instead be thought of as a set of tools.
Each particpant may start the workshop with different tools.

In order to get a better sense of what topics in R we should focus more heavily on, let's do a quick assessment.
The results will help us shape instruction so that we can ensure we're meeting everyone's needs.

**Instructions:** 

Answer the following 5 questions to the best of your knowledge and write down your answers.

### Question 1

read csv

### Question 2

What does the following expression return?

`max(abs(c(-5, 1, 5)))`
```{r q1, echo = FALSE}
question("",
         answer("-5"),
         answer("1", correct = T),
         answer("3", correct = T),
         answer("An error"),
         allow_retry = T)

```

Choose one:

- A. -5
- B. 1
- C. 5
- D. An error

Your answer:

### Question 3

If x and y are both data.frames defined by:

```{r}
x <- data.frame(z = 1:2)
y <- data.frame(z = 3)
```

which of the following expressions would be a correct way to combine them into one data.frame that looks like this:

```
z
-
1
2
3
```

(i.e. one column with the numbers 1, 2, and 3 in it)

Choose one or more:

- A. `join(x, y)`
- B. `c(x, y)`
- C. `rbind(x, y)`
- D. `x + y`

### Question 4
  
What is the output of the following code chunk?

```{r, eval = FALSE}
x <- data.frame(x = 1:10, y = 1:10)
dim(x)
```

- A. 2
- B. 10
- C. 2 10
- D. 10 2
- E. 10 10


### Question 5
  
Given the following data.frame,

x <- data.frame(y = 1:10)

Which expression(s) return a `data.frame` with rows where y is greater than 5 (i.e. 6 - 10)

Choose one or more:

- A. `x[x$y > 5,]`
- B. ``x$y > 5`
- C. `x[which(x$y > 6),]`
- D. `x[y > 5,]`
- E. `subset(x, y > 5)`


### Question 6
  
Given the following data.frame `iris`,

```{r, echo = FALSE}
head(iris)
```

Which expression return a `data.frame` with rows where Species is "setosa" and only the columns `Sepal.Length` and `Species`?

- A. `iris["setosa", c("Sepal.Length", "Species")]`
- B. `iris[Species == "setosa" & c("Sepal.Length", "Species"),]`
- C. `subset(iris, Species == "setosa", c("Sepal.Length", "Species"))`
- D. `iris[Species == "setosa", c("Sepal.Length", "Species")]`
- E. `iris["setosa", c(1,5)]`


For the questions below, unless otherwise specified, select the output of the following code chunks.

### Question 7

```{r, eval = FALSE}
x <- 2
x ^ 2
```

- A. 2
- B. "2 ^ 2"
- C. 4
- D. "x ^ 2"
- E. 8


### Question 8

```{r, eval = FALSE}
x <- 1; y <- 2; x + y;
```

- A. 1
- B. 2
- C. 3
- D. Error: object 'y' not found
- E. 1 2

### Question 9 

```{r, eval = FALSE}
x <- "hello"
y <- "world"
paste(x, y)
```

- A. "hello world"
- B. "hello" "world"
- C. "hello, world"
- D. "helloworld"
- E. "hello;world"

### Question 10

```{r, eval = FALSE}
x <- list(1, 2, 3)
y <- list(4, 5, 6)
z <- c(x, y)
length(z)
```

- A. 2
- B. 3
- C. 6
- D. 21
- E. None of the above

### Question 11

```{r, eval = FALSE}
x <- data.frame(x = 1:6)
y <- data.frame(x = 1:7)
z <- rbind(x, y)
nrow(z)
```

- A. 1
- B. 6
- C. 7
- D. 13
- E. 14

### Question 12

```{r, eval = FALSE}
x <- NA

if (is.na(x)) {
  print("foo")
} else {
  print("bar")
}
```

- A. "foo"
- B. "bar"
- C. NA
- D. "foobar"
- E. TRUE

### Question 13

What will the following code print?

```{r, eval = FALSE}
numbers <- seq(1, 10)

for (number in numbers) {
  if (number %% 2){
    print(number)
  }
}
```

- A. The even numbers from 0 to 10 (including 0 and 10)
- B. The even numbers from 0 to 10 (not including 0 and 10)
- C. The odd numbers from 0 to 10 (including 1 and 9)
- D. The odd numbers from 0 to 10 (not including 1 and 9)
- E. None of the above


### Question 14

How many values of `x` will the following code print?
```{r, eval = FALSE}
x <- 10

while (x >= 0) {
  print(x)
  
  if (x == 5) {
    break
  }
  
  x <- x - 1
}
```

- A. 1
- B. 4
- C. 5
- D. 6
- E. 10


### Question 15 

```{r, eval = FALSE}
x <- c(1, "2", 3)
class(x)
```

- A. numeric
- B. character
- C. factor
- D. "numeric" "character" "numeric"
- E. NA

### Question 16

```{r, eval = FALSE}
x <- list(c(1,2), 2, 3)
y <- lapply(x, cumsum)
y[[3]] - y[[1]]
```

- A. 0
- B. 3 
- C. 2 0
- D. 0 0
- E. 2

### Question 17

```{r, eval = FALSE}
x <- data.frame(letter = LETTERS)
class(x[1,])
```

- A. "data.frame"
- B. "numeric"
- C. "matrix"
- D. "vector"
- E. "factor"

### Question 18

```{r, eval = FALSE}
x <- data.frame(letter = LETTERS)
class(x[1, 1, drop = FALSE])
```

- A. "data.frame"
- B. "numeric"
- C. "matrix"
- D. "vector"
- E. "factor"

### Question 19

```{r, eval= FALSE}
x <- data.frame(x = 1)
y <- data.frame(x = 2)
z <- rbind(x[1,1], y[1,1])
class(z)
```

- A. "data.frame"
- B. "numeric"
- C. "matrix"
- D. "vector"
- E. "factor"

### Question 20

```{r, eval = FALSE}
rep(TRUE, 5) & rep(FALSE, 5)
```

- A. FALSE
- B. TRUE
- C. FALSE FALSE FALSE FALSE FALSE
- D. TRUE TRUE TRUE TRUE TRUE
- E. TRUE TRUE TRUE TRUE TRUE FALSE FALSE FALSE FALSE FALSE

### Question 21

```{r, eval = FALSE}
rep(TRUE, 5) && rep(FALSE, 5)
```

- A. FALSE
- B. TRUE
- C. FALSE FALSE FALSE FALSE FALSE
- D. TRUE TRUE TRUE TRUE TRUE
- E. TRUE TRUE TRUE TRUE TRUE FALSE FALSE FALSE FALSE FALSE

### Question 22

```{r, eval = FALSE}
x <- c(1, 2, 3)
y <- c("A", "B", "C")
z <- c(x, y)
class(z)
```

- A. numeric
- B. character
- C. factor
- D. "numeric" "numeric" "numeric" "character" "character" "character"
- E. NA

### Question 23

```{r, eval = FALSE}
x <- c(1, 2, NA, 4, NA)
length(is.na(x))
```

- A. 2
- B. FALSE FALSE TRUE FALSE TRUE
- C. 5
- D. NA
- E. FALSE

### Question 24

```{r ex25, eval = FALSE}
x <- c(1, NA, 3)
y <- c(NA, 2, NA)

z <- data.frame(x,y)

sum(is.na(z))
```

- A. NA NA NA
- B. 1 2 3
- C. TRUE TRUE TRUE
- D. FALSE FALSE FALSE
- E. TRUE
    
  
### Question 25

```{r, eval = FALSE}
x <- c(1, "A", 3)
as.numeric(x)
```

- A. NA NA NA
- B. 1 A 3
- C. 1 NA 3
- D. TRUE FALSE TRUE
- E. FALSE
    
    
    
    
## R overview

Based on previous R training: https://github.nceas.ucsb.edu/Training/R-intro.
Instructor will go over these live with the classroom, running 

- Basic syntax
  - Variables & assignemnt
  
### The assignment operator, `<-`

One of the things we'll do all the time is save some value to a variable.
Here, we save the word "apple" to a variable called `fruit`

```{r}
fruit <- "apple"
fruit
```

Notice the last line with just `fruit` on it.
Typing just the variable name in just prints the value to the Console.

R has a flexible syntax.
The following two lines of code are identical to the above one.

```{r}
fruit<-"apple"
fruit    <-     "apple"
```

### R as a calculator: `+ - * / > >= %% %/%` etc
  
```{r}
2+2
2 * 3
2 ^ 3
5/2
```

Comparison:

```{r}
2 == 1
2 == 2
3 > 2
2 < 3 # Same as above
"apple" == "apple"
"apple" == "pair"
"pair" == "apple" # Order doesn't matter for ==
```


### Types of variables

#### Vectors

When we run a line of code like this:

```{r}
x <- 2
```

We're assigning 2 to a variable `x`.
`x` is a variable but it is also a "numeric vector" of length 1.

```{r}
class(x)
length(x)
```

Above, we ran two function: `class` and `length` on our variable `x`.
Running functions is a very common thing you'll do in R.
Every function has a name, following by a pair of `()` with something inside.

We can make a numeric vector that is longer like so:

```{r}
x <- c(1, 2, 3) # Use the `c` function to put things together
```

Notice we can also re-define a variable at a later point just like we did above.

```{r}
class(x)
length(x)
```

R can store much more than just numbers though.
Let's start with strings of characters, which we've already seen:

```{r}
fruit <- "apple"
class(fruit)
length(fruit)
```

Depending on your background, you may be surprised that the result of running `length(fruit)` is 1 because "apple" is five characters long.

It turns out that `fruit` is a character vector of length one, just like our numeric vector from before.
To find out the number of characters in "apple", we have to use another function:

```{r}
nchar(fruit)
nchar("apple")
```

Let's make a character vector of more than length one and take a look at how it works:

```{r}
fruits <- c("apple", "banana", "strawberry")
length(fruits)
nchar(fruits)
fruits[1]
```

Smushing character vectors together can be done with `paste`:

```{r}
paste("key", "lime", "pie")
```

#### Lists

Vectors and lists look similar in R sometimes but they have very different uses:

```{r}
c(1, "apple", 3)
list(1, "apple", 3)
```

#### data.frames
Most of the time when doing analysis in R you will be working with `data.frames`.
`data.frames` are tabular, with column headings and rows of data, just like a CSV file.

We create new `data.frames` with a relevantly-named function:

```{r}
mydata <- data.frame(site = c("A", "B", "C"),
                     temp = c(20, 30, 40))
mydata
```

Or we can read in a CSV from the file system and turn it into a `data.frame` in order to work with it in R:

```{r}
mydata <- read.csv("data.csv")
mydata
```

We can find out how many rows of data `mydata` has in it:

```{r}
nrow(mydata)
```

We can return just one of the columns:

```{r}
mydata$type
unique(mydata$type)
```
sort

If we want to sort `mydata`, we use the `order` function (in kind of a weird way):

```{r}
mydata[order(mydata$type),]
```

Let's break the above command down a bit.
We can access the individual cells of a `data.frame` with a new syntax element: `[` and `[`:

```{r}
mydata[1,] # First row
mydata[,1] # First column
mydata[1,1] # First row, first column
mydata[c(1,5),] # First and second row
mydata$type # Column named 'type'
```

So what does that `order` function do?

```{r}
?order # How to get help in R!
order(c(1, 2, 3))
order(c(3, 2, 1))
order(mydata$type)
```

So `order(mydata$type)` is returning the rows of `mydata`, by row number, in sorted order.

We can also return just certain rows, based upon criteria:

```{r}
mydata[mydata$type == "fruit",]
mydata$type == "fruit"
```

In this case, instead of indexing the rows by number, we're using TRUEs and FALSEs.

**Exercise:** Subset `mydata` to the vegetables instead of the fruit

```{r}
# Your code here
```

Another handy way to subset `data.frame` is with the `subset` function:
    
```{r}
subset(mydata, type == "fruit") # Equivalent to mydata[mydata$type == "fruit",]
```

There are a lot of useful functions to help us work with `data.frame`s:

```{r}
str(mydata)
summary(mydata)
```

Our `data.frame`s won't always be so small as this example one.
Let's look at a larger one:

```{r}
library(ggplot2)
data("diamonds")
diamonds
```

**Exercise:** How many rows does diamonds have in it? How many columns?

We can look at the first few rows with `head`, just like on the command line:

```{r}
head(diamonds)
```

or the last few:

```{r}
tail(diamonds)
```
So far, this has probably been a bit boring.
Let's do something interesting and also something that R is very good at: Plotting and modeling!

Let's plot the relationship between diamond price and carat:

```{r}
plot(price ~ carat, data = diamonds)
```

The above syntax, `price ~ carat` uses a `response ~ predictor` form or `y ~ x`.

We can also fit a linear model to the same relationship:

```{r}
mod <- lm(price ~ carat, data = diamonds)
```

Above, we saved our linear model to a variable named `mod` so we can use it later.
We can look at the result of model fitting with `summary`.

```{r}
summary(mod)
```

And we can also plot the line of best fit on the scatterplot:

```{r}
plot(price ~ carat, data = diamonds)
abline(mod$coefficients[[1]], mod$coefficients[[2]], col = "red", lwd = 5)
```

### More advanced topics to cover if time allows:

- Modeling
  - `sample` / `runif` / `rnorm`
    - sampling a data.frame
- Common util functions
    - `table`
- Iteration
    - `for` loops
        - When do we use for loops?
    - `while` loops
- apply-family
  - `sapply`
  - `lapply`
- functions: We'll skip this as it will be taught in another lesson
- Writing faster code
  - What is 'vectorization'?
  - Don't grow variables
- Debugging
  - traceback()
  - browser()
  - breakpoints in RStudio
  

## Summary

By the end of this lesson, you should have feel touched up on your general R skills and you also should have seen some of the trickier parts of R.
Hopefully having seen the trickier parts of R will help later on down the road.

## Resources

Other good resources:

- One of our lessons on R: https://github.nceas.ucsb.edu/Training/R-intro
- Great, free book: http://r4ds.had.co.nz/
- DataCamp's R intro, https://www.datacamp.com/courses/free-introduction-to-r
Sources:
- Used https://www.rstudio.com/resources/cheatsheets/ to make sure I wasn't missing any basic stuff